\pkg{evtree}: Evolutionary Learning of Globally Optimal Classification and Regression Trees in \proglang{R}

### 2011-08-05 pkg/evtree/inst/doc/evtree.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@197 255f4249-2263-4417-b1ea-8498779e27a7
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@198 255f4249-2263-4417-b1ea-8498779e27a7
started some touch-ups
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@205 255f4249-2263-4417-b1ea-8498779e27a7

\documentclass[nojss]{jss}
\usepackage{amssymb, amsmath, amsthm} 
%\usepackage{longtable}
\usepackage{booktabs}
\usepackage{listings}
\raggedbottom

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% almost as usual
\author{Thomas Grubinger\\ Innsbruck Medical University 
\And Achim Zeileis\\ Universit\"at Innsbruck 
\And Karl-Peter Pfeiffer\\ Innsbruck Medical University
}

\title{\pkg{evtree}: Evolutionary Learning of Globally Optimal Classification and Regression Trees in \proglang{R}}
%% for pretty printing and a nice hypersummary also set:
\Plainauthor{Thomas Grubinger, Achim Zeileis and Karl-Peter Pfeiffer} %% comma-separated
\Plaintitle{evtree: Evolutionary Learning of Globally Optimal Classification and Regression Trees in R} %% without formatting
\Shorttitle{Evolutionary Learning of Globally Optimal Classification and Regression Trees in R} %% a short title (if necessary)

%% an abstract and keywords
\Abstract{
Commonly used classification and regression tree methods like the CART algorithm are recursive partitioning methods that build the model in a forward stepwise search. Although this approach is known to be an efficient heuristic, the results of recursive tree methods are only locally optimal, as splits are chosen to maximize homogeneity at the next step only. An alternative way to search over the parameter space of trees is to use global optimization methods like evolutionary algorithms. This paper describes the \pkg{evtree} package, which implements an evolutionary algorithm for learning globally optimal classification and regression trees in \proglang{R}. CPU and memory-intensive tasks are fully computed in \proglang{C++} while the \pkg{partykit} \citep{hothorn2009partykit} package is leveraged to represent the resulting trees in \proglang{R}, providing unified infrastructure for summaries, visualizations, and predictions. \code{evtree} is compared to \code{rpart} \citep{therneau1997introduction}, the open-source CART implementation, and \code{ctree} \citep{hothorn2006urp} (conditional inference trees). On 20 evaluated data-sets, including several benchmark problems from the UCI machine learning repository, \code{evtree} models offered at least similar and most of the time increased predictive accuracy.  
}

\Keywords{machine learning, classification trees, regression trees, evolutionary algorithms, \proglang{R}}
\Plainkeywords{machine learning, classification trees, regression trees, evolutionary algorithms, R} %% without formatting
%% at least one keyword must be supplied

%% publication informatione
%% NOTE: Typically, this can be left commented and will be filled out by the technical editor
%% \Volume{13}
%% \Issue{9}
%% \Month{September}
%% \Year{2004}
%% \Submitdate{2004-09-29}
%% \Acceptdate{2004-09-29}

%% The address of (at least) one author should be given
%% in the following format:
\Address{
  Thomas Grubinger, Karl-Peter Pfeiffer\\
  Innsbruck Medical University\\
  Department for Medical Statistics, Informatics and Health Economics\\
  6020, Innsbruck\\
  E-mail: \email{thomas.grubinger@i-med.ac.at}\\
  URL: \url{http://www.i-med.ac.at/msig/mitarbeiter/grubinger}\\

  Achim Zeileis \\
  Universit\"at Innsbruck\\
  Department of Statistics\\
  6020, Innsbruck\\
  E-mail: \email{Achim.Zeileis@R-project.org}\\
  URL: \url{http://eeecon.uibk.ac.at/~zeileis/}
}

%% It is also possible to add a telephone and fax number
%% before the e-mail in the following format:
%% Telephone: +43/1/31336-5053
%% Fax: +43/1/31336-734

%% for those who use Sweave please include the following line (with % symbols):
%% need no \usepackage{Sweave.sty}

%% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

A very simple example that illustrates the limitation of recursive partitioning methods is depicted in Figure~\ref{fig:illustration}. The example only contains two independent variables and can be solved with only three splits that partition the input space into four regions. As expected the recursive partitioning methods \code{rpart} and \code{ctree} fail to find any split at all, as the impurity of the resulting subsets cannot be reduced by the first split. For methods that explore $\Theta$ in a more global fashion it is straightforward to find an optimal solution to this problem. One solution, which is the tree constructed by \code{evtree} is given by:

<< echo = FALSE, split= TRUE>>=
library("evtree")
X1 <- c(rep(0.25, 4), rep(0.75, 4), rep(1.25 ,4),  rep(1.75, 4))
X2 <- c(rep(seq(0.25, 1.75 ,0.5),4))
Y <- array(1,16)
Y[X1 < 1 & X2 < 1]=2
Y[X1 > 1 & X2 > 1]=2
chess22 <- as.data.frame(cbind(X1, X2, Y))
chess22[,3] <- as.factor( chess22[,3] )
levels(chess22[,3]) = c("class: o","class: x")
print(evtree(Y~.,data=chess22, minbucket = 1, minsplit=2))
@

\setkeys{Gin}{width=0.4\textwidth}
\begin{figure}[tbh!]
    \centering
<<fig = TRUE, echo = FALSE>>=
par(cex.axis=1.5, cex.lab=1.5)
plot(chess22[chess22[,3]=="class: o",1], chess22[chess22[,3]=="class: o",2], type="p",  pch= c("x"), col=colors()[320], xlab="X1", ylab="X2", xlim=c(0,2), ylim=c(0,2), cex=2)
points(chess22[chess22[,3]=="class: x",1], chess22[chess22[,3]=="class: x",2], pch= c("o"), cex=2)
par(cex.axis=1.5, cex.lab=1.5)
@
\caption{Class distribution of  the $(X_1, X_2)$-plane. The two classes are indicated by black $(o)$ and grey $(x)$.}
\label{fig:illustration}
\end{figure}

@@

\code{evtree} provides an efficient implementation of an evolutionary algorithm that builds classification trees in \proglang{R}. CPU- and memory- intensive tasks are fully computed in \proglang{C++}, while the user interfaces and plot functions are written in \proglang{R}. The \code{.C()} interface \citep{chambers2008software} was used to pass arguments between the two languages. \code{evtree} depends on the \pkg{partykit} package  \citep{hothorn2009partykit}, which provides an infrastructure for representing, summarizing, and visualizing tree-structured models. The supplied functionality includes standard \code{print()}, \code{summary()}, and \code{plot()} functions to display trees and a \code{predict()} function to predict the class labels of new data. The arguments of  \code{evtree()} are

\begin{verbatim}
evtree(formula, data = list(), weights = NULL, subset = NULL, control = 
evtree.control(...), ...)
\end{verbatim}

where \code{formula} and \code{data} are a standard \proglang{R} formula of type \code{y ~ x1 + x2} with the respective data. \code{weights} is an optional vector specifying weights to be used in the fitting process. The option vector subset specifies a subset of observations to be used in the fitting process. \code{control} comprises a list of control parameters

\begin{verbatim}
evtree.control(minbucket = 7L, minsplit = 20L, maxdepth = 9L, 
niterations = 10000L, ntrees = 100L, alpha = 1, operatorprob = list(
pmutatemajor = 0.2, pmutateminor = 0.2, pcrossover = 0.2, psplit = 0.2, 
pprune = 0.2), seed = NULL, ...)
\end{verbatim}

where the parameters \code{minbucket}, \code{minsplit}, and \code{maxdepth} constrain the solution to a minimum number of observations in each terminal node, a minimum number of observation in each internal node and a maximum tree depth. Note that the memory requirements increase by the square of the maximum tree depth. Parameter \code{alpha} regulates the complexity part of the evaluation function and refers to $\alpha$ in Equation~\ref{eq:classification} for classification trees and Equation~\ref{eq:regression} for regression trees. \code{niterations} and \code{ntrees} specify the maximum number of iterations and the number of trees in the population. With the argument \code{operatorprob}, user-specified probabilities for the variation operators can be defined. Argument \code{seed} is an optional seed for the random number generator, which allows models to be reproduced by re-using the same seed. If not specified, the random number generator is initialized by the system time. In this case, iterative calls of \code{evtree} will possibly result in different models.

<<>>=
library("evtree")
data("iris")
model <- evtree(Species ~ ., iris, control = evtree.control(seed = 1000))
@
The fitted model can be examined using the \code{print()} and \code{plot()} functions.

<<>>=
print(model)
@

\setkeys{Gin}{width=1\textwidth}
\begin{figure}[tbh!]
    \begin{center}
<<echo= FALSE>>=
par(cex.axis=1.5, cex.lab=1.5)
<<fig=TRUE, echo=TRUE, width = 13.5, height= 8.5>>=
plot(model)
@
     \end{center}
     \caption{Example of a tree plotted by the \code{plot()} function}
     \label{fig:example}
\end{figure}

The resulting plot is displayed in Figure~\ref{fig:example}. \code{predict()} can be used for the prediction of the response.

<<>>=
predict(model, newdata = iris[c(10, 35, 60, 85, 110, 135), -5])
@

@@

%% vorberechnungen
<<label=figciplots, echo = FALSE>>=
rm(list=ls())
library("plotrix")
library("lme4")
library("multcomp")
load(file="./results/breastcancer.RData")
load(file="./results/glass.RData")
load(file="./results/credit.RData")
load(file="./results/heart.RData")
load(file="./results/pima.RData")
load(file="./results/contraceptive.RData")
load(file="./results/dna.RData")
load(file="./results/musk.RData")
load(file="./results/spam.RData")
load(file="./results/vowel.RData")
load(file="./results/ionosphere.RData")
load(file="./results/magicgamma.RData")
load(file="./results/hdg0202.RData")
load(file="./results/hdg0502.RData")
load(file="./results/mel0101.RData")
load(file="./results/bostonhousing.RData")
load(file="./results/servo.RData")

@@

\setkeys{Gin}{width=1\textwidth}
\begin{figure}[tbh!]
    \centering
<<fig = TRUE, echo = FALSE, width=14, height=7.9>>=
plotd("rpart")
@
\caption{Performance comparison of \code{evtree} and \code{rpart}. Prediction error is compared by the relative difference of the misclassification rate or the mean-squared error. The complexity is compared by the relative difference of the number of terminal nodes.}
 \label{fig:crpart}
\end{figure}

\begin{figure}[tbh!]
    \centering
<<fig = TRUE, echo = FALSE, width=14, height=7.9>>=
plotd("ctree")
@
\caption{Performance comparison of \code{evtree} and \code{ctree}. Prediction error is compared by the relative difference of the misclassification rate or the mean-squared error. The complexity is compared by the relative difference of the number of terminal nodes.}
\label{fig:cptree}
\end{figure}

@@

\setkeys{Gin}{width=0.4\textwidth}
\begin{figure}[tbh!]
    \centering
<<fig = TRUE, echo = FALSE>>=
par(cex.axis=1.5, cex.lab=1.5)
ch <- chessboard44(totalpoints= 2000)
par(cex.axis=1.5, cex.lab=1.5)
plot(ch[ch[,9]==1,1], ch[ch[,9]==1,2], type="p",  pch= c("x"), col=colors()[320], xlab="X1", ylab="X2", cex=1)
points(ch[ch[,9]==0,1], ch[ch[,9]==0,2], pch= c("o"))
@
\caption{Class distribution of the simulated $4 \times 4$ chess-board problem plotted on the $(X_1, X_2)$-plane. The two classes are indicated by black $(o)$ and grey $(x)$.}
\label{fig:chess}
\end{figure}

@@

<<label =  tab:resultsChessboard,  echo = FALSE, results=tex>>=
library(xtable)
load(file = "./results/chessboard44_0.RData")
load(file = "./results/chessboard44_5.RData")
load(file = "./results/chessboard44_10.RData")

@@

\bibliography{article.bib}

>>>>

\documentclass[nojss]{jss}
\usepackage{amssymb, amsmath, amsthm, booktabs, thumbpdf}

%% Sweave/vignette information and metadata
%% need no \usepackage{Sweave}
\SweaveOpts{engine = R, eps = FALSE, keep.source = TRUE}
%\VignetteIndexEntry{Evolutionary Learning of Globally Optimal Classification and Regression Trees in R}
%\VignetteDepends{stats,evtree,partykit,mlbench,plotrix,lme4,multcomp,rpart,ctree,xtable}
%\VignetteKeywords{machine learning, classification trees, regression trees, evolutionary algorithms, R}
%\VignettePackage{evtree}

\author{Thomas Grubinger\\ Innsbruck Medical University 
   \And Achim Zeileis\\ Universit\"at Innsbruck 
   \And Karl-Peter Pfeiffer\\ Innsbruck Medical University
}
\Plainauthor{Thomas Grubinger, Achim Zeileis, Karl-Peter Pfeiffer}

\title{\pkg{evtree}: Evolutionary Learning of Globally Optimal Classification and Regression Trees in \proglang{R}}
\Plaintitle{evtree: Evolutionary Learning of Globally Optimal Classification and Regression Trees in R}
\Shorttitle{\pkg{evtree}: Evolutionary Learning of Globally Optimal Trees in \proglang{R}}

\Abstract{
  Commonly used classification and regression tree methods like the CART
  algorithm are recursive partitioning methods that build the model in a forward
  stepwise search. Although this approach is known to be an efficient heuristic,
  the results of recursive tree methods are only locally optimal, as splits are
  chosen to maximize homogeneity at the next step only. An alternative way to
  search over the parameter space of trees is to use global optimization methods
  like evolutionary algorithms. This paper describes the \pkg{evtree} package,
  which implements an evolutionary algorithm for learning globally optimal
  classification and regression trees in \proglang{R}. CPU and memory-intensive
  tasks are fully computed in \proglang{C++} while the \pkg{partykit}
  \citep{hothorn2009partykit} package is leveraged to represent the resulting
  trees in \proglang{R}, providing unified infrastructure for summaries,
  visualizations, and predictions. \code{evtree} is compared to \code{rpart}
  \citep{therneau1997introduction}, the open-source CART implementation, and
  \code{ctree} \citep{hothorn2006urp} (conditional inference trees). On 20
  evaluated data-sets, including several benchmark problems from the UCI machine
  learning repository, \code{evtree} models offered at least similar and most of
  the time increased predictive accuracy.  
}

\Keywords{machine learning, classification trees, regression trees, evolutionary algorithms, \proglang{R}}
\Plainkeywords{machine learning, classification trees, regression trees, evolutionary algorithms, R}

\Address{
  Thomas Grubinger, Karl-Peter Pfeiffer\\
  Innsbruck Medical University\\
  Department for Medical Statistics, Informatics and Health Economics\\
  6020, Innsbruck, Austria\\
  E-mail: \email{thomas.grubinger@i-med.ac.at}\\
  URL: \url{http://www.i-med.ac.at/msig/mitarbeiter/grubinger}\\

  Achim Zeileis \\
  Universit\"at Innsbruck\\
  Department of Statistics\\
  6020, Innsbruck, Austria\\
  E-mail: \email{Achim.Zeileis@R-project.org}\\
  URL: \url{http://eeecon.uibk.ac.at/~zeileis/}
}

@@

A very simple example that illustrates the limitation of recursive partitioning methods is depicted in Figure~\ref{fig:illustration}. The example only contains two independent variables and can be solved with only three splits that partition the input space into four regions. As expected the recursive partitioning methods \code{rpart} and \code{ctree} fail to find any split at all, as the impurity of the resulting subsets cannot be reduced by the first split. For methods that explore $\Theta$ in a more global fashion it is straightforward to find an optimal solution to this problem. One solution, which is the tree constructed by \code{evtree} is given by: 
%
<< echo = FALSE, split= TRUE>>=
library("evtree")
X1 <- c(rep(0.25, 4), rep(0.75, 4), rep(1.25 ,4),  rep(1.75, 4))
X2 <- c(rep(seq(0.25, 1.75 ,0.5),4))
Y <- array(1,16)
Y[X1 < 1 & X2 < 1]=2
Y[X1 > 1 & X2 > 1]=2
chess22 <- as.data.frame(cbind(X1, X2, Y))
chess22[,3] <- as.factor( chess22[,3] )
levels(chess22[,3]) = c("class: o","class: x")
print(evtree(Y~.,data=chess22, minbucket = 1, minsplit=2))
@

\setkeys{Gin}{width=0.4\textwidth}
\begin{figure}[tbh!]
\centering
<<fig = TRUE, echo = FALSE>>=
par(cex.axis=1.5, cex.lab=1.5)
plot(chess22[chess22[,3]=="class: o",1], chess22[chess22[,3]=="class: o",2], type="p",  pch= c("x"), col=colors()[320], xlab="X1", ylab="X2", xlim=c(0,2), ylim=c(0,2), cex=2)
points(chess22[chess22[,3]=="class: x",1], chess22[chess22[,3]=="class: x",2], pch= c("o"), cex=2)
par(cex.axis=1.5, cex.lab=1.5)
@
\caption{\label{fig:illustration} Class distribution of  the $(X_1, X_2)$-plane. The two classes are indicated by black $(o)$ and grey $(x)$.}
\end{figure}

@@

\code{evtree} provides an efficient implementation of an evolutionary algorithm that builds classification trees in \proglang{R}. CPU- and memory- intensive tasks are fully computed in \proglang{C++}, while the user interfaces and plot functions are written in \proglang{R}. The \code{.C()} interface \citep{chambers2008software} was used to pass arguments between the two languages. \code{evtree} depends on the \pkg{partykit} package  \citep{hothorn2009partykit}, which provides an infrastructure for representing, summarizing, and visualizing tree-structured models. The supplied functionality includes standard \code{print()}, \code{summary()}, and \code{plot()} functions to display trees and a \code{predict()} function to predict the class labels of new data. The arguments of  \code{evtree()} are
%
\begin{Code}
  evtree(formula, data = list(), weights = NULL, subset = NULL,
    control = evtree.control(...), ...)
\end{Code}
%
where \code{formula} and \code{data} are a standard \proglang{R} formula of type \code{y ~ x1 + x2} with the respective data. \code{weights} is an optional vector specifying weights to be used in the fitting process. The option vector subset specifies a subset of observations to be used in the fitting process. \code{control} comprises a list of control parameters
%
\begin{Code}
  evtree.control(minbucket = 7L, minsplit = 20L, maxdepth = 9L, 
    niterations = 10000L, ntrees = 100L, alpha = 1,
    operatorprob = list(pmutatemajor = 0.2, pmutateminor = 0.2,
      pcrossover = 0.2, psplit = 0.2, pprune = 0.2),
    seed = NULL, ...)
\end{Code}
%
where the parameters \code{minbucket}, \code{minsplit}, and \code{maxdepth} constrain the solution to a minimum number of observations in each terminal node, a minimum number of observation in each internal node and a maximum tree depth. Note that the memory requirements increase by the square of the maximum tree depth. Parameter \code{alpha} regulates the complexity part of the evaluation function and refers to $\alpha$ in Equation~\ref{eq:classification} for classification trees and Equation~\ref{eq:regression} for regression trees. \code{niterations} and \code{ntrees} specify the maximum number of iterations and the number of trees in the population. With the argument \code{operatorprob}, user-specified probabilities for the variation operators can be defined. Argument \code{seed} is an optional seed for the random number generator, which allows models to be reproduced by re-using the same seed. If not specified, the random number generator is initialized by the system time. In this case, iterative calls of \code{evtree} will possibly result in different models.
%
<<>>=
library("evtree")
data("iris")
model <- evtree(Species ~ ., iris, control = evtree.control(seed = 1000))
@
%
The fitted model can be examined using the \code{print()} and \code{plot()} functions. 
%
<<>>=
print(model)
@

\setkeys{Gin}{width=1\textwidth}
\begin{figure}[t!]
\centering
<<echo= FALSE>>=
par(cex.axis=1.5, cex.lab=1.5)
<<fig=TRUE, echo=TRUE, width = 13.5, height= 8.5>>=
plot(model)
@
\caption{\label{fig:example} Example of a tree plotted by the \code{plot()} function.}
\end{figure}

The resulting plot is displayed in Figure~\ref{fig:example}. \code{predict()} can be used for the prediction of the response.
%
<<>>=
predict(model, newdata = iris[c(10, 35, 60, 85, 110, 135), -5])
@

@@

%% vorberechnungen
<<label=figciplots, echo = FALSE>>=
rm(list = ls())
library("plotrix")
library("lme4")
library("multcomp")
load(file="./results/breastcancer.RData")
load(file="./results/glass.RData")
load(file="./results/credit.RData")
load(file="./results/heart.RData")
load(file="./results/pima.RData")
load(file="./results/contraceptive.RData")
load(file="./results/dna.RData")
load(file="./results/musk.RData")
load(file="./results/spam.RData")
load(file="./results/vowel.RData")
load(file="./results/ionosphere.RData")
load(file="./results/magicgamma.RData")
load(file="./results/hdg0202.RData")
load(file="./results/hdg0502.RData")
load(file="./results/mel0101.RData")
load(file="./results/bostonhousing.RData")
load(file="./results/servo.RData")

@@

\setkeys{Gin}{width=1\textwidth}
\begin{figure}[t!]
\centering
<<fig = TRUE, echo = FALSE, width=14, height=7.9>>=
plotd("rpart")
@
\caption{\label{fig:crpart} Performance comparison of \code{evtree} and \code{rpart}. Prediction error is compared by the relative difference of the misclassification rate or the mean-squared error. The complexity is compared by the relative difference of the number of terminal nodes.}
\end{figure}

\begin{figure}[tbh!]
\centering
<<fig = TRUE, echo = FALSE, width=14, height=7.9>>=
plotd("ctree")
@
\caption{\label{fig:cptree} Performance comparison of \code{evtree} and \code{ctree}. Prediction error is compared by the relative difference of the misclassification rate or the mean-squared error. The complexity is compared by the relative difference of the number of terminal nodes.}
\end{figure}

@@

\setkeys{Gin}{width=0.4\textwidth}
\begin{figure}[tbh!]
\centering
<<fig = TRUE, echo = FALSE>>=
par(cex.axis=1.5, cex.lab=1.5)
ch <- chessboard44(totalpoints= 2000)
par(cex.axis=1.5, cex.lab=1.5)
plot(ch[ch[,9]==1,1], ch[ch[,9]==1,2], type="p",  pch= c("x"), col=colors()[320], xlab="X1", ylab="X2", cex=1)
points(ch[ch[,9]==0,1], ch[ch[,9]==0,2], pch= c("o"))
@
\caption{\label{fig:chess} Class distribution of the simulated $4 \times 4$ chess-board problem plotted on the $(X_1, X_2)$-plane. The two classes are indicated by black $(o)$ and grey $(x)$.}
\end{figure}

@@

<<label =  tab:resultsChessboard,  echo = FALSE, results=tex>>=
library("xtable")
load(file = "./results/chessboard44_0.RData")
load(file = "./results/chessboard44_5.RData")
load(file = "./results/chessboard44_10.RData")

@@

\bibliography{evtree}
